Formal models of biological networks in ANIMO 2

### 2013-12-06 Paper.tex
Added more about model checking, trying to adapt better the 'performance' tests

\title{Building formal models of biological networks avoiding formal training}

@@

\section*{Background}\label{sec:introduction}
In living cells, processes are regulated by networks of interacting molecules.
Aberrations in these networks underlie a wide range of pathologies. The development of new therapies
requires a thorough insight in the functioning of these networks. Obtaining such insight can be a challenging task.
Feedback loops and crosstalk between pathways lead to an intricate wiring of the network.
Hence, it is necessary to study the ensemble of molecules involved, because the 
behaviour of individual molecules is not sufficient for a complete understanding. 
Since the human brain is ill-suited to grasp the non-linear dynamics of these complex networks and
the entailed emergent properties, the role of computational support is increasing
in molecular biology.

@@

\tas\ have been developed to model the dynamic behaviour of systems with processes running in parallel~\cite{timed-automata-alur-dill}. 
As such, \tas\ have been applied in communication protocols and industrial control engineering~\cite{ta-audio-protocol,ta-wap-gateway,ta-ws-bap}. The 
parallels between these application areas and regulatory processes in cells have triggered the step towards 
their use in biology.
\cite{ta-siebert} use \tas\ to extend a classical modelling paradigm~\cite{thomas-formalism},
allowing to add temporal dynamics to gene network models.
\cite{bartocci-oscillators} describe a model of biological oscillators and test 
synchronization properties in this dynamic system.
A discretization of ODEs to \tas\ is proposed by~\cite{oded-ode-ta-discretization}, applying
a translation between the two formalisms to an example gene regulatory network. Two 
different approaches to transforming
a Petri net model into \tas\ are presented by~\cite{ta-giapponesi},
who also address the important issue of state space explosion in their paper.
Finally, \cite{ta-radiazioni} propose an \emph{ad hoc} \tas\ model of a radiation treatment
system, which is then validated through the analysis tool UPPAAL~\cite{uppaal}.\\
Each of these approaches has been successfully validated, demonstrating the potential of \tas\
in biological applications. However, these approaches were all limited to simple
or specific examples and none of these modelling methods
has led to a tool implementation of the proposed method to encourage a broader use
of \tas\ in molecular biology. We have recently introduced the software tool ANIMO
(Analysis of Networks with Interactive MOdelling,~\cite{animo-ieee}), which aims
at making available the power of \tas\ to the experimental biologists without requiring
them to be fluent in any mathematical formalism.

@@

@@

The output of the ANIMO model in Figure~\ref{fig:drosophila-model} closely matches the original ODE model.
In particular, the oscillations in both models show the same periods and phases (see Suppl. Fig~\ref{suppl-fig:grafici-drosophila}).
Due to the compositional nature of \tas\, ANIMO allows for intuitive \emph{in silico} knock-out experiments,
by right-clicking a node in the model and disabling it. Such experiments have been done
before~\cite{drosophila-ode-model} and give similar results in our model.
% Details of the comparison
% between the ANIMO model and the original ODE model is given in Supplementary Section~\ref{suppl-sec:animo-drosophila}.

@@

\subsection*{Exploiting formal methods: model checking}\label{subsec:model-checking}
To show why one should use a formal model like \tas, we demonstrate the application
one of the biggest advantages of \tas: state space-based analyses such as
model checking~\cite{model-checking} provided by UPPAAL~\cite{uppaal}.

The technique of model checking consists in verifying the truth of some properties
that describe the behaviour of a model. This is usually done by exploring every
possible evolution of the model starting from an initial state. The set of all states
that can be reached by a model is called \emph{state space}, and formulae describing
the evolution of a system are usually described by (a subset of) computation tree
logic (CTL,~\cite{ctl}).

% \subsection*{Reaction-centered VS reactant-centered}\label{sec:animo-comparison}
% We will now apply some basic model checking queries to the case study presented in~\cite{animo-bibe},
% measuring the performances of the two modeling approaches. This will allow us to
% evaluate the benefit brought by the shift in perspective from a reaction- to a reactant-centered model.

The case study we use as a testbed is the network model shown in the \emph{Network} panel in Figure~\ref{fig:cytoscape},
which represents signaling events downstream of growth factors EGF (epidermal growth factor)
and NGF (nerve growth factor) in PC12 cells (a cell line used to study neuronal differentiation).
The model topology proposed in~\cite{egf-ngf} was analyzed
with an ANIMO model based on the reaction-centered approach, reproducing the experimentally 
observed ERK (extracellular signal-regulated kinase) activity changes~\cite{animo-bibe}.
In particular, a 10 minutes stimulation with EGF resulted in transient behavior (i.e. peak-shaped,
see also the graph in Fig.~\ref{fig:cytoscape}), while NGF stimulation led to sustained activity (see Fig.~\ref{fig:comparison-graph-animo}).

This case study will also be used to evaulate the improvement in performances introduced
by the new version of ANIMO, which is described in the Methods section.

\subsubsection*{Simulation-based analysis}
We start by evaluating the cost in terms of time of simulations. This is a particularly
important aspect to consider, as during the model building phase a user may need to perform a relatively
large number of simulations,
continuously adapting the topology or quantitative parameters of a network model. In order to make
the modeling approach in ANIMO as interactive as possible, it is desirable to decrease dead times,
and this translates into reducing the computational cost of model analysis as much as possible.
UPPAAL's statistical model checking engine~\cite{uppaal-smc} is based on the generation of random walks (i.e., simulation runs) from a \tas\ model.
In this experiment, we query UPPAAL for simulation runs on the models
generated by ANIMO applying the reaction- and reactant-centered modeling approaches to the case study.
To define the initial state of the model, we consider the starting condition to be the treatment with 50~ng/ml NGF, which translates into
setting the activity level of node NGF to 15/15, while changing EGF to be at 0/15 activity. We have chosen this configuration as the
model is expected to continue reacting indefinitely (sustained ERK activity), without reaching a deadlock state where all reactions are inactive.
This makes the model more demanding when generating the state space
because having more reactions occur means having more states in the transition system underlying the \tas\ model.
Table~\ref{tab:sim-100} illustrates the computation time and memory usage when performing 100 simulation runs on each of the
two considered models. Computing the simulation runs took about $92 \%$ less time with the reactant-centered model, using $87 \%$
less memory. This decrease in computation time for long simulation runs brings the approach nearer to the idea
of interactive exploration of a network.

\subsubsection*{Model checking}
Next, we set out to test the model checking performances using the old \tas\ model used in ANIMO and
the new one introduced in ANIMO 2. We will compare the execution times and memory requirements for a number of interesting queries:
\begin{itemize}
  \item (1) and (2): {\tt A[] not deadlock}. The model continues to execute indefinitely ({\tt A} refers to all
      possible paths in the transition system of the model, and {\tt []} asks the property to always hold along a path).
  \item (3): {\tt RKIP $<$ 10 --$>$ ERK $>=$ 40}. After RKIP (Raf kinase inhibitory protein) activity has been lowered, ERK activity increases. As in the model RKIP
      has 20 levels of granularity and ERK has 100 levels, {\tt RKIP $<$ 10} means that RKIP is less than half active, and
      {\tt ERK $>=$ 40} means that ERK activity is at least $40 \%$.
  \item (4): {\tt E<> RKIP $<$ 10}. Find a point when RKIP is low ({\tt E} asks for the existence of at least one path
      for which the property holds, while {\tt <>} requires the property to hold at least once along a given path).
      This query is expected to generate a trace, the last point of which will be used as initial configuration for model checking queries (5) and (6).
  \item (5): {\tt A[] ERK $<$ 70} and (6): {\tt A[] ERK $>$ 35}. Once RKIP activity has significantly decreased, ERK activity is sustained at an intermediate level.
\end{itemize}
The initial conditions are:
\begin{itemize}
  \item (1): EGF = 15/15 and NGF = 0/15, all others as the original configuration, corresponding to the treatment condition with
	    100 ng/ml EGF.
  \item (2) - (4): EGF = 0/15, NGF = 15/15, all others as the original configuration, corresponding to the treatment condition with
	    50 ng/ml NGF\footnote{In the laboratory experimental setting, NGF is used at a lower concentration than EGF, but it is still enough to saturate all NGF receptors, which
	    are rarer than EGF receptors.}.
  \item (5) and (6): all activities as in the last state of the trace computed from query~(4).
\end{itemize}

@@

>>>>

\title{Formal models of biological networks in ANIMO 2} %Building formal models of biological networks avoiding formal training}

@@

\section*{Background}\label{sec:introduction}
\subsection*{Modelling in cell biology}
In living cells, processes are regulated by networks of interacting molecules.
Aberrations in these networks underlie a wide range of pathologies. The development of new therapies
requires a thorough insight in the functioning of these networks. Obtaining such insight can be a challenging task.
Feedback loops and crosstalk between pathways lead to an intricate wiring of the network.
Hence, it is necessary to study the ensemble of molecules involved, because the 
behaviour of individual molecules is not sufficient for a complete understanding. 
Since the human brain is ill-suited to grasp the non-linear dynamics of these complex networks and
the entailed emergent properties, the role of computational support is increasing
in molecular biology.

@@

\subsection*{\tas\ for biological models}
\tas\ have been developed to model the dynamic behaviour of systems with processes running in parallel~\cite{timed-automata-alur-dill}. 
As such, \tas\ have been applied in communication protocols and industrial control engineering~\cite{ta-audio-protocol,ta-wap-gateway,ta-ws-bap}. The 
parallels between these application areas and regulatory processes in cells have triggered the step towards 
their use in biology.
\cite{ta-siebert} use \tas\ to extend a classical modelling paradigm~\cite{thomas-formalism},
allowing to add temporal dynamics to gene network models.
\cite{bartocci-oscillators} describe a model of biological oscillators and test 
synchronization properties in this dynamic system.
A discretization of ODEs to \tas\ is proposed by~\cite{oded-ode-ta-discretization}, applying
a translation between the two formalisms to an example gene regulatory network. Two 
different approaches to transforming
a Petri net model into \tas\ are presented by~\cite{ta-giapponesi},
who also address the important issue of state space explosion in their paper.
Finally, \cite{ta-radiazioni} propose an \emph{ad hoc} \tas\ model of a radiation treatment
system, which is then validated through the analysis tool UPPAAL~\cite{uppaal}.\\
Each of these approaches has been successfully validated, demonstrating the potential of \tas\
in biological applications. However, these approaches were all limited to simple
or specific examples and none of these modelling methods
has led to a tool implementation of the proposed method to encourage a broader use
of \tas\ in molecular biology. We have recently introduced the software tool ANIMO
(Analysis of Networks with Interactive MOdelling,~\cite{animo-ieee}), which aims
at making available the power of \tas\ to the experimental biologists without requiring
them to be fluent in any mathematical formalism.

@@

\subsection*{A brief introduction to ANIMO}
{\Large [[briefly describe ANIMO, its modelling assumptions and its intended use]]}

@@

The output of the ANIMO model in Figure~\ref{fig:drosophila-model} closely matches the original ODE model.
In particular, the oscillations in both models show the same periods and phases (see Suppl. Fig~\ref{suppl-fig:grafici-drosophila}).
Due to the compositional nature of \tas\, ANIMO allows for intuitive \emph{in silico} knock-out experiments,
by right-clicking a node in the model and disabling it. Such experiments have been done
before~\cite{drosophila-ode-model} and give similar results in our model.
% Details of the comparison
% between the ANIMO model and the original ODE model is given in Supplementary Section~\ref{suppl-sec:animo-drosophila}.
\\
{\Large [[Insert description of K.O. experiments with Drosophila model]]}

@@

\subsection*{Testing the performances of ANIMO 2.0}\label{subsec:model-checking}
% \subsection*{Reaction-centered VS reactant-centered}\label{sec:animo-comparison}
% We will now apply some basic model checking queries to the case study presented in~\cite{animo-bibe},
% measuring the performances of the two modeling approaches. This will allow us to
% evaluate the benefit brought by the shift in perspective from a reaction- to a reactant-centered model.
In addition to some new features, the new ANIMO 2.0 provides a boost in terms of performances
(see Methods section for details).
In order to measure how significant the performance increase is, we applied a number of
ANIMO analysis queries to a case study, comparing the performances of ANIMO 2.0 with
the previous version of the tool. Moreover, ANIMO 2.0 provides the biolgist with
a proper access to the formal technique of model checking~\cite{model-checking}. As we will show, the
performance improvement is even more significant in this case, increasing
the potential for interactivity also in this area.

The case study we use as a testbed is the network model shown in the \emph{Network} panel in Figure~\ref{fig:cytoscape},
which represents signaling events downstream of growth factors EGF (epidermal growth factor)
and NGF (nerve growth factor) in PC12 cells (a cell line used to study neuronal differentiation).
The topology for this network was proposed in~\cite{egf-ngf}, and it was analyzed
with an ANIMO model based on the reaction-centered approach, reproducing the experimentally 
observed ERK (extracellular signal-regulated kinase) activity changes~\cite{animo-bibe}.
In particular, a 10 minutes stimulation with EGF resulted in transient behavior (i.e. peak-shaped,
see also the graph in Fig.~\ref{fig:cytoscape}), while NGF stimulation led to sustained activity (see Fig.~\ref{fig:comparison-graph-animo}).

\subsubsection*{Simulation-based analysis}
The most basic analysis that can be performed with ANIMO is based on simulating the behaviour
of the network during a given period of time. As this type of analysis is also normally
used when looking for a match between model parameters and experimental data, the computation
speed of a simulation run deeply affects the level of interactivity perceived by the user.
In order to make
the modeling approach in ANIMO as smooth as possible, it is desirable to decrease dead times,
i.e. significantly reduce simulation costs.
% UPPAAL is used in ANIMO to obtain simulation runs from a \tas\ model. While that process is
% very fast, 
% In this experiment, we query UPPAAL for simulation runs on the models
Our first experiment compares the execution time for simulation runs on the \tas\ models
generated by ANIMO applying the reaction- and reactant-centered modeling approaches to the case study.
To define the initial state of the model, we consider the starting condition to be the treatment with 50~ng/ml NGF, which translates into
setting the activity level of node NGF to 15/15, while changing EGF to be at 0/15 activity. We have chosen this configuration as the
model is expected to continue reacting indefinitely (sustained ERK activity), without reaching a deadlock state where all reactions are inactive.
This makes the model more demanding when generating the state space
because having more reactions occur means having more states in the transition system underlying the \tas\ model.
Table~\ref{tab:sim-100} illustrates the computation time and memory usage when performing 100 simulation runs on each of the
two considered models. Computing the simulation runs took about $92 \%$ less time with the reactant-centered model, using $87 \%$
less memory. This decrease in computation time for long simulation runs brings the approach nearer to the idea
of interactive exploration of a network.

\subsubsection*{A good reason to use formal methods: model checking}
% Next, we set out to test the model checking performances using the old \tas\ model used in ANIMO and
% the new one introduced in ANIMO 2.
To show why one should use a formal model like \tas, we demonstrate the application
one of the biggest advantages of \tas: model checking~\cite{model-checking}.

The technique of model checking consists in verifying the truth of some properties
that describe the behaviour of a model. This is usually done by exploring every
possible execution path of the model starting from its initial configuration. The set of all
the configurations that can be reached by a model during all its possible executions
is called \emph{state space}. The properties on the evolution of the model are
usually expressed in a formal language such as computation tree logic (CTL,~\cite{ctl}),
which allows to formally describe how and when a property should hold. Examples of such
properties can be expressed in natural language as: ``the activity of node A
stays always above 20'', ``it is possible to reach a state where nodes A and B are both
active for at least 60 \%'', ``whenever node A becomes active for at least 20 \%, node
B becomes active for at least 30 \%''.

Model checking performances in ANIMO 2.0 have been tested using the case study of EGF-NGF signalling as reference.
We compared the execution times and memory requirements in ANIMO and ANIMO 2.0 for a number of interesting queries:
\begin{itemize}
  \item (1) and (2): {\tt A[] not deadlock}. The model continues to execute indefinitely ({\tt A} refers to all
      possible paths in the transition system of the model, and {\tt []} asks the property to always hold along a path).
  \item (3): {\tt RKIP $<$ 10 --$>$ ERK $>=$ 40}. After RKIP (Raf kinase inhibitory protein) activity has been lowered, ERK activity increases. As in the model RKIP
      has 20 levels of granularity and ERK has 100 levels, {\tt RKIP $<$ 10} means that RKIP is less than half active, and
      {\tt ERK $>=$ 40} means that ERK activity is at least $40 \%$.
  \item (4): {\tt E<> RKIP $<$ 10}. Find a point when RKIP is low ({\tt E} asks for the existence of at least one path
      for which the property holds, while {\tt <>} requires the property to hold at least once along a given path).
      This query is expected to generate a trace, the last point of which will be used as initial configuration for model checking queries (5) and (6).
  \item (5): {\tt A[] ERK $<$ 70} and (6): {\tt A[] ERK $>$ 35}. Once RKIP activity has significantly decreased, ERK activity is sustained at an intermediate level.
\end{itemize}
The initial conditions are:
\begin{itemize}
  \item (1): EGF = 15/15 and NGF = 0/15, all others as the original configuration, corresponding to the treatment condition with
	    100 ng/ml EGF.
  \item (2) - (4): EGF = 0/15, NGF = 15/15, all others as the original configuration, corresponding to the treatment condition with
	    50 ng/ml NGF\footnote{In the laboratory experimental setting, NGF is used at a lower concentration than EGF, but it is still enough to saturate all NGF receptors, which
	    are rarer than EGF receptors.}.
  \item (5) and (6): all activities as in the last state of the trace computed from query~(4).
\end{itemize}

@@

\subsection*{Model checking for bigger models}
When a model becomes particularly complex, with many events occurring concurrently,
the state space of the model tends to \emph{grow exponentially} in size, dramatically
increasing the time requested for the analysis of bigger models. To avoid the so-called
state space explosion, an approach that is becoming popular: the so-called \emph{statistical
model checking}.
